Physics of Tidal Dissipation: Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar Envelopes and Massive Stellar Binaries

### 2019-12-22 writeups/Su_IGW_break.tex
check in draft 191222

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\mathbf{#1}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

\title[Internal Gravity Wave Breaking in White Dwarf Binaries]{Internal Gravity
Wave Breaking in White Dwarf Binaries}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai,$^1$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
}

@@

\begin{document}\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\renewcommand*{\sectionautorefname}{Section}
\maketitle

% cp ../sims/agg.png plots; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/f_amps.png plots/lin_amps.png; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/fluxes.png plots/lin_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps.png plots/nl_f_amps.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/fluxes.png plots/nl_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_refl.png plots/nl_f_refl.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_ri.png plots/nl_f_ri.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/front.png plots/nl_front.png;

@@

\section{Introduction}

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA) \citep{lisa}. They are also thought to
produce interesting optical transients such as underluminous supernovae
\citep{underlum}, Ca-rich fast transients \citep{carich}, and tidal novae
\citep{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g. \citealp{Ia0,webbink} or more recently
\citealp{Ia1,Ia2}). While presently only a few tens of compact WD binaries are
known \citep{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds \citep{lsst_wd} \citep[results based on
\emph{Gaia}'s second data release have already begun to
appear][]{gaiaDD,gaiaDD2}, and the Large Synoptic Survey Telescope (LSST, first
light scheduled for 2020) will likely detect a few thousand more
\citep{lsst_wd}. These observations will significantly advance the understanding
of WD binaries and their evolution.

@@

These two zones are separated by a \emph{critical layer}. The majority of our
paper is dedicated to characterizing the behavior of the critical layer when
interacting with a continuous train of IGW excited from the bottom of the fluid.
IGW are generally \emph{anti}-diffusive, in that they steepen shear flows
\citep{lindzen_qbo,lecoanet_meanflow}, and act to narrow the critical layer. We
find this steepening is suppressed by the Kelvin-Helmholtz instability and
turbulence within the very narrow critical layer. By careful accounting of the
momentum flux budget about the critical layer, we are able to model the
propagation of the critical layer, reflection of the incident IGW, and other
momentum redistribution behaviors. Our results are numerically converged
for sufficiently large resolutions.

In \autoref{s:equations}, we will describe the system of equations we will use
to analyze IGW breaking. In \autoref{s:theory}, we review existing understanding
of wave breaking and present new analytical results. In \autoref{s:numerics}, we
describe our numerical setup, which in \autoref{s:weak_sim} we validate in the
weak forcing limit against linear theory. In \autoref{s:sim}, we present the
results of simulation of IGW breaking and our characterization of the critical
layer. Finally, we summarize and conclude in \autoref{s:discussion}.

\section{Problem Description}\label{s:equations}

We consider a incompressible, isothermal fluid, representative of degenerate
matter in WDs. We use barotropic equation of state $P(\rho, T) = P(\rho)$ as a
first approximation. As we are interested in dynamics far from the center of the
WD, we approximate the gravitational field as uniform. We model the background
density stratification as $\overline{\rho} = \overline{\rho}_0 e^{-z/H}$ for
some reference density $\overline{\rho}_0$ (we generally notate background
quantities with overbars and perturbation quantities with primes). We
consider a 2D fluid with for computational feasibility. While
it is well-known wave breaking is a 3D process \citep{klostermeyer,winters1994},
the dynamical effect of the breaking process is likely to be similar in 2D
\citep{barker_ogilvie}. Finally, we simplify by considering simple
plane-parallel IGW\@. Since wave breaking generally occurs near the surface of
WDs \citep[$\gtrsim 0.9R_{WD}$][]{fullerI}, the curvature of the wave front is
negligible.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p{\bm{u} \cdot \bm{\nabla}}$ is the Lagrangian or
material derivative, and $\bm{u}, \rho, P$ denote the velocity field, density
and pressure respectively. We denote $-g\uv{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\bm{\nabla}\overline{P} = -\overline{\rho} g\uv{z}$ and so $\overline{P} =
\overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\uv{x}$ is permitted at hydrostatic equilibrium, but we will
assume no background flow, so $\bm{u} = \bm{u}'$. Physically, this assumption
corresponds to a non-rotating star, or going to the corotating frame of a
rigidly rotating star.

In practice, it is convenient to introduce the coordinate $\Upsilon = \ln
\frac{\rho}{\overline{\rho}}$ \citep[e.g.][]{lecoanet_anel}. This both
identically enforces $\rho > 0$ and eliminates the stiff term
$\frac{\bm{\nabla}P}{\rho}$. We also define reduced pressure $\varpi =
\frac{P}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0\label{eq:nl_upsilon_u}.
    \end{align}
\end{subequations}
In the new coordinates, hydrostatic equilibrium corresponds to $\Upsilon = 0,
\overline{\varpi} = gH$.

@@

\subsection{Linear Analysis}

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \bm{u} \cdot
\bm{\nabla}$, we may linearize \autoref{se:nl_upsilon}. The solution in this
linear regime is given by \citep{drazin,sutherland0}:
\begin{equation}
    u_z\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:k0z_sign}
\end{equation}
where $A$ is the amplitude and $\omega$ satisfies dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed $c_s \to \infty$, in
which the \emph{Brunt-V\"ais\"al\"a frequency}
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$ \citep[see
e.g.][]{sutherland0}.

In the short-wavelength/WKB limit $\abs{k_{z}H} \gg 1$, the solution exhibits
the following characteristics:
\begin{itemize}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation is always violated for sufficiently large $z$.

    \item The phase and group velocities are respectively:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}},\\
            \bm{c}_{g} &= N\frac{\p{k_{z}^2 + \frac{1}{4H^2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\p{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\p{ \p{k_{z}H}^{-2} }
        \approx 0$. In the Boussinesq approximation where density stratification
        outside of $N^2$ is completely neglected, the phase and group velocity
        are exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

    \item The averaged horizontal momentum flux $F$ is
        \begin{align}
            F &\equiv \ev{\rho u_{x} u_{z}}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}u_{z}\;\mathrm{d}x,
                    \label{eq:F_def}\\
                &\approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{align}
        Thus, indeed $F > 0$ for an upward propagating IGW $c_{g, z} > 0$.
\end{itemize}

@@

To model continuous excitation of IGWs deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the WD, and
downwards, which are not physically meaningful, but are damped away by the
damping layers described in \autoref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ localized around
$z_0$ with small width $\sigma$ by replacing \autoref{eq:nl_up_density} with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the given
$k_{x}$, $\omega(k_{x}, k_{z})$ will propagate.

In the linearized system, the effect of this forcing can be solved exactly
(\autoref{s:force_solved}). If we approximate $k_zH \gg 1, \sigma \ll H$, the
solution can be approximated as two plane waves propagating away from the
forcing zone
\begin{align}
    u_{z}&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        e^{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain excited deep
in the atmosphere.

@@

Above, we neglected advective terms $\pd{}{t} \gg \bm{u} \cdot \bm{\nabla}$.
However, since $\abs{\bm{u}} \propto e^{z/2H}$ in an infinite domain all IGWs
eventually grow to nonlinear amplitudes and break. Nonlinear wave breaking is
expected to be important in WDs \citep{fullerI,fullerII}. We can
estimate the height of wave breaking using $\frac{\abs{\bm{u}}
\abs{\bm{k}}}{\omega} \sim 1$. This can be rewritten using the Lagrangian
displacement $-i\omega \bm{\xi} = \bm{u}$:
\begin{equation}
    \xi_z k_z \gtrsim 1.\label{eq:nl}
\end{equation}

% An equivalent estimate given by \citep{eliassen_palm_cite,sutherland0} is that
% IGWs break when the wave-induced mean flow is equal to the horizontal phase
% velocity
% \begin{equation}
%     \overline{U}(z) = \frac{\ev{u_{x}u_{z}}_x}{c_{g,z}} \sim
%         \frac{\omega}{k_x},\label{eq:rip}
% \end{equation}
% where \overline{U}(z) \equiv \ev{u_x}_x$ is the mean flow of the fluid.
% \autoref{eq:rip} provides a leading-order estimate for \overline{U}(z)$ which
% may further evolve over time (see \autoref{ss:crit_layer}); for a general mean
% flow, \autoref{eq:rip} also defines the criterion for a \emph{critical
% layer}. This wave-induced mean flow is analogous to Stokes' drift for surface
% waves and can be derived by considering the propagation of $F$
% (\autoref{eq:F_def}) into a medium at rest. Evaluating \autoref{eq:rip}
% yields agreement with \autoref{eq:nl}.

\citealp{drazin, klostermeyer, winters1994} describe the onset of wave breaking.
At intermediate amplitudes, wave breaking occurs via triadic resonances which
transfer energy from the ``parent’’ internal gravity wave (in our problem, the
tidally forced wave) to ``daughter’’ internal gravity waves on smaller length
scales that efficiently damp. The horizontal momentum flux decreases from $F$ to
$0$ over this breaking region. The lost flux is deposited into a horizontal mean
flow
\begin{equation}
    \overline{U}(z, t) \equiv \ev{u_x}_x.\label{eq:mean_flow}
\end{equation}
As the mean flow grows, a \emph{critical layer} may form, discussed below.

@@

A horizontal shear flow $\overline{U}(z, t)\hat{x}$ enters the fluid equations
via the Lagrangian derivative, which can be decomposed as
\begin{equation}
    \md{}{t} = \pd{}{t} + \overline{U} \pd{}{x} + \p{\bm{u}' \cdot \vec{\nabla}},
\end{equation}
where $\bm{u}'$ is the velocity field \emph{without} the shear flow, for
clarity. Thus, $\overline{U}$ has the effect of Doppler shifting the time
derivative into the frame comoving with the mean flow. If $\overline{U}$ is
roughly constant, then the behavior of a small plane wave perturbation satisfies
modified dispersion relation
\begin{equation}
    \p{\omega - \overline{U}k_x}^2 =
        \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \frac{1}{4H^2}}.
        \label{eq:disp_rel_U}
\end{equation}
This is just \autoref{eq:disp_rel} with $\omega \to \omega - \overline{U}k_x$.
It is apparent that if $\overline{U} = \overline{U}_c$, where
\begin{equation}
    \overline{U}_c \equiv \frac{\omega}{k_x},
\end{equation}
then the dispersion relation is singular and the linear solution breaks down.
Physically, this corresponds to the Doppler-shifted frequency of the IGW being
zero. Anywhere $\overline{U} = \overline{U}_c$ is called a \emph{critical
layer}.

>>>>

    \newcommand*{\scinot}[2]{#1\times10^{#2}}
    \newcommand*{\rd}[2]{\frac{\mathrm{d}#1}{\mathrm{d}#2}}
    \newcommand*{\rtd}[2]{\frac{\mathrm{d}^2#1}{\mathrm{d}#2^2}}
    \newcommand*{\pd}[2]{\frac{\partial#1}{\partial#2}}
    \newcommand*{\ptd}[2]{\frac{\partial^2#1}{\partial#2^2}}
    % inline
    \newcommand*{\mdil}[2]{\mathrm{D}#1/\mathrm{D}#2}
    \newcommand*{\pdil}[2]{\partial#1/\partial#2}
    \newcommand*{\rdil}[2]{\mathrm{d}#1/\mathrm{d}#2}
    \newcommand*{\md}[2]{\frac{\mathrm{D}#1}{\mathrm{D}#2}}
    \newcommand*{\at}[1]{\left.#1\right|}
    \newcommand*{\abs}[1]{\left|#1\right|}
    \newcommand*{\ev}[1]{\left\langle#1\right\rangle}
    \newcommand*{\p}[1]{\left(#1\right)}
    \newcommand*{\s}[1]{\left[#1\right]}
    \newcommand*{\z}[1]{\left\{#1\right\}}
    \newcommand*{\bm}[1]{\mathbf{#1}}
    \newcommand*{\uv}[1]{\hat{\mathbf{#1}}}
    \DeclareMathOperator*{\argmin}{argmin}
    \DeclareMathOperator*{\argmax}{argmax}
    \DeclareMathOperator*{\med}{med}
    \DeclareMathOperator*{\erf}{erf}

\title[Internal Gravity Wave Breaking]{Physics of Tidal Dissipation:
Hydrodynamical Simulations of Internal Gravity Wave Breaking in Stellar
Envelopes and Massive Stellar Binaries}
\author[Y. Su et\ al.]{
Yubo Su,$^1$,
Daniel Lecoanet,$^2$
Dong Lai,$^1$
\\
$^1$ Cornell Center for Astrophysics and Planetary Science, Department of
Astronomy, Cornell University, Ithaca, NY 14853, USA
\\
$^2$ Princeton Center for Theoretical Science, Princeton University, Princeton,
NJ 08544, USA
}

@@

\begin{document}\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% cp ../sims/agg.png plots; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/f_amps.png plots/lin_amps.png; cp ../sims/2d_4_fourier/snapshots_lin_0_masked/fluxes.png plots/lin_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps2.png plots/nl_f_amps2.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_amps.png plots/nl_f_amps.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/fluxes.png plots/nl_fluxes.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_refl.png plots/nl_f_refl.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/f_ri.png plots/nl_f_ri.png; cp ../sims/2d_4_fourier/snapshots_yubo_nu1_vhres/front.png plots/nl_front.png;

@@

\section{Introduction}\label{s:intro}

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA) \citep{lisa}. They are also thought to
produce interesting optical transients such as underluminous supernovae
\citep{underlum}, Ca-rich fast transients \citep{carich}, and tidal novae
\citep{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g. \citet{Ia0,webbink} or more recently
\citet{Ia1,Ia2}). While presently only a few tens of compact WD binaries are
known \citep{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds \citep{lsst_wd} \citep[results based on
\emph{Gaia}'s second data release have already begun to
appear][]{gaiaDD,gaiaDD2}, and the Large Synoptic Survey Telescope (LSST, first
light scheduled for 2020) will likely detect a few thousand more
\citep{lsst_wd}. These observations will significantly advance the understanding
of WD binaries and their evolution.

@@

These two zones are separated by a \emph{critical layer}. The interaction of IGW
and critical layers plays an important role in the tidal synchronization process
of stellar binaries \citep{zahn75,gn89} as well as in terrestrial phenomena such
as the quasi-biennial oscillation \citep{lindzen_qbo}. The majority of our paper
is dedicated to characterizing the behavior of the critical layer when
interacting with a continuous train of IGW excited from the bottom of the fluid.
IGW are generally \emph{anti}-diffusive, in that they steepen shear flows
\citep{lindzen_qbo,lecoanet_meanflow}, and act to narrow the critical layer. We
find this steepening is suppressed by the Kelvin-Helmholtz instability and
turbulence within the very narrow critical layer. By careful accounting of the
momentum flux budget about the critical layer, we are able to model the
propagation of the critical layer, reflection of the incident IGW, and other
momentum redistribution behaviors. Our results are numerically converged for
sufficiently large resolutions.

In Section~\ref{s:equations}, we will describe the system of equations we will
use to analyze IGW breaking. In Section~\ref{s:theory}, we review existing
understanding of wave breaking and present new analytical results. In
Section~\ref{s:numerics}, we describe our numerical setup, which in
Section~\ref{s:weak_sim} we validate in the weak forcing limit against linear
theory. In Section~\ref{s:sim}, we present the results of simulation of IGW
breaking and our characterization of the critical layer. Finally, we summarize
and conclude in Section~\ref{s:discussion}.

\section{Problem Setup}\label{s:equations}

We consider a stratified, incompressible, isothermal fluid representing a
stellar envelope or atmosphere. We use barotropic equation of state $P(\rho, T)
= P(\rho)$ as a first approximation. We study dynamics in 2D, so that fluid
variables depend only on the Cartesian coordinates $x$ and $z$. While it is
well-known wave breaking is a 3D process \citep{klostermeyer,winters1994}, the
dynamical effect of the breaking process is likely to be similar in 2D
\citep{barker_ogilvie}. We approximate the gravitational field as uniform,
pointing in the $(-\uv{z})$ direction. The plane-parallel approximation is
justified since wave breaking generally occurs near the stellar surface. The
background density stratification is given by
\begin{equation}
    \overline{\rho} = \overline{\rho}_0 e^{-z/H},
\end{equation}
with $\overline{\rho}_0$ some reference density (we generally denote background
quantities with overbars and perturbation quantities with primes).

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \bm{\nabla} \cdot \bm{u} &= 0,\label{eq:nl_incomp}\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\bm{u}}{t} + \frac{\bm{\nabla}P}{\rho} + g\uv{z} &=
            0\label{eq:nl_mom},
    \end{align}
\end{subequations}
where $\mdil{}{t} = \pdil{}{t}\,+\,\p{\bm{u} \cdot \bm{\nabla}}$ is the
Lagrangian or material derivative, and $\bm{u}, \rho, P$ denote the velocity
field, density and pressure respectively. We denote $-g\uv{z}$ constant
gravitational acceleration. Note that hydrostatic equilibrium ($\pdil{}{t} = 0$)
of the background implies $\bm{\nabla}\overline{P} = -\overline{\rho} g\uv{z}$,
so that $\overline{P} = \overline{\rho} gH$. A vertically-stratified shear flow
$\overline{u}_x(z)\uv{x}$ is permitted in hydrostatic equilibrium, but we will
assume no such background flow, so $\bm{u} = \bm{u}'$. Physically, this
assumption corresponds to a non-rotating star, or going to the corotating frame
of a rigidly rotating star.

For convenience, we introduce the dimensionless density variable $\Upsilon$ and
the reduced pressure $\varpi$ \citep[e.g.][]{lecoanet_anel} via
\begin{align}
    \Upsilon &\equiv \ln \frac{\rho}{\bar{\rho}},\\
    \varpi &\equiv \frac{P}{\rho}.
\end{align}
These variables automatically enforce $\rho > 0$ and eliminate the stiff term
$\vec{\nabla} P / \rho$ in the Euler equation. In terms of $\Upsilon$ and
$\varpi$, the second two equations in~\eqref{se:nl_orig} become
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \overline{\rho}}{z} &= 0
            ,\label{eq:nl_up_density} \\
        \md{\bm{u}}{t} + \bm{\nabla}\varpi + \varpi\bm{\nabla}\Upsilon
            - \frac{\varpi}{H}\uv{z} + g\uv{z} &= 0\label{eq:nl_upsilon_u}.
    \end{align}
\end{subequations}
Hydrostatic equilibrium corresponds to $\Upsilon = 0, \overline{\varpi} = gH$.

@@

\subsection{Linear Analysis}\label{ss:lin_analysis}

In the small perturbation limit, we may linearize
Eqs.~\eqref{se:nl_upsilon}. The solution in this linear regime is given by
\citep{drazin,sutherland0}
\begin{equation}
    u_z'\p{x, z, t} = Ae^{z/2H}\cos\p{k_{x}x + k_{z}z - \omega t},
        \label{eq:lin_sol}
\end{equation}
where $A$ is a constant amplitude, and the frequency $\omega$ and the wave
number $\p{k_x, k_z}$ satisfy the dispersion relation
\begin{equation}
    \omega^2 = \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel}
\end{equation}
Our equations are valid in the limit of large sound speed ($c_s \to \infty$), in
which the \emph{Brunt-V\"ais\"al\"a frequency} $N$ is given by
\begin{equation}
    N^2 \equiv g^2\p{\rd{\rho}{P} - \frac{1}{c_s^2}} = \frac{g}{H},
\end{equation}
is constant. Other dynamical quantities are simply related to $u'_z$.

In the short-wavelength/WKB limit ($\abs{k_{z}H} \gg 1$), the solution exhibits
the following characteristics:
\begin{enumerate}
    \item The amplitude of the wave grows with $z$ as $e^{z/2H}$. Thus, the
        linear approximation always breaks down for sufficiently large $z$.

    \item The phase and group velocities are given by:
        \begin{align}
            \bm{c}_{p} &=
                \p{k_{x}\uv{x} + k_{z}\uv{z}}\frac{\omega}
                {k_{x}^2 + k_{z}^2 + \p{2H}^{-2}},\\
            \bm{c}_{g} &= N\frac{\s{k_{z}^2 + \p{2H}^{-2}}\uv{x}
                - \p{k_{x}k_{z}\uv{z}}}
                {\s{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}^{3/2}}.\label{eq:vg}
        \end{align}
        We note $\bm{c}_{p} \cdot \bm{c}_g = \mathcal{O}\s{\p{k_{z}H}^{-2}}
        \approx 0$. In the Boussinesq approximation where terms
        $\mathcal{O}\p{H^{-2}}$ are ignored, the phase and group velocities are
        exactly orthogonal \citep{drazin,sutherland1}. We use the convention
        where upward propagating IGW have $c_{g, z} > 0$, $k_z < 0, k_x > 0$.

    \item The averaged horizontal momentum flux $F$ (in the $+\uv{z}$
        direction) carried by the IGW is defined by
        \begin{equation}
            F(z, t) \equiv \ev{\rho u_{x}' u_{z}'}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_{x}'u_{z}'\;\mathrm{d}x.
                    \label{eq:F_def}
        \end{equation}
        The notation $\ev{\dots}_x$ denotes averaging over the $x$ domain. For
        the linear solution (Eq.~\eqref{eq:lin_sol}), this evaluates to
        \begin{equation}
            F \approx -\frac{A^2}{2}\overline{\rho}_0\frac{k_{z}}{k_{x}},
                    \label{eq:S_lin}
        \end{equation}
        Thus, indeed $F > 0$ for an upward propagating IGW ($c_{g, z} > 0$).
\end{enumerate}

@@

To model continuous excitation of IGWs deep in the stellar envelope propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain. Our forcing excites both IGWs
propagating upwards, imitating a wave tidally excited deeper in the star, and
downwards, which are not physically meaningful, but are dissipated by a
damping zone described in Section~\ref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. We implement forcing with strength $C$ localized around
$z_0$ with small width $\sigma$ by replacing Eq.~\eqref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_{z}\pd{\ln \overline{\rho}}{z}
        = Ce^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p{k_{x}x - \omega t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $z$ power spectrum, but only the
$k_{z}$ satisfying dispersion relation (Eq.~\eqref{eq:disp_rel}) for the given
$k_{x}$ and $\omega$ will propagate.

In the linearized system, the effect of this forcing can be solved exactly
(see Appendix~\ref{s:force_solved}). If we approximate $\abs{k_zH} \gg 1, \sigma
\ll H$, the solution can be approximated as two plane waves propagating away
from the forcing zone
\begin{align}
    u_{z}&(x, z, t) \approx{} \frac{C}{2k_z}\frac{gk_x^2}{\omega^2}
        \exp\p{-\frac{k_z^2\sigma^2}{2}}
        \sqrt{2\pi \sigma^2} \nonumber\\
        &{}\times\begin{cases}
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x + k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & \text{for }z > z_0,\\[5pt]
        e^{\frac{z - z_0}{2H}}\sin\p{k_{x}x - k_{z}(z - z_0) - \omega t
            + \frac{k_z\sigma^2}{2H}}
            & \text{for }z < z_0.\\
    \end{cases}\label{eq:uz_lin}
\end{align}
The $z > z_0$ region models an upward propagating IGW wavetrain. The $x$
component of the velocity can be obtained by the incompressibility constraint
(Eq.~\eqref{eq:nl_incomp}).

@@

As the upward propagating IGW grows in amplitude ($\abs{\bm{u}} \propto
e^{z/2H}$), it is expected to break due to nonlinear effects. We can estimate
the height of wave breaking as when $\abs{\bm{u}} \sim \omega / \abs{\bm{k}}$.
This can be rewritten using the Lagrangian displacement $\bm{\xi} = \bm{u} /
\p{-i\omega}$:
\begin{equation}
    \abs{\xi_z k_z} \gtrsim 1.\label{eq:nl}
\end{equation}

\citet{drazin, klostermeyer, winters1994} describe the onset of wave breaking in
some detail. At intermediate amplitudes, wave breaking occurs via triadic
resonances which transfer energy from the ``parent’’ IGW to ``daughter’’ waves
on smaller length scales that efficiently damp. The horizontal momentum flux
decreases from $F$ to $0$ over this breaking region. The lost flux is deposited
into a horizontal mean flow
\begin{equation}
    \overline{U}(z, t) \equiv \ev{u_x}_x.\label{eq:mean_flow}
\end{equation}
As the mean flow grows, a \emph{critical layer} may form, as discussed below.

@@

A horizontal shear flow $\overline{U}(z, t)\uv{x}$ enters the fluid equations
via the Lagrangian derivative, which can be decomposed as
\begin{equation}
    \md{}{t} = \pd{}{t} + \overline{U} \pd{}{x} + \p{\bm{u}' \cdot \vec{\nabla}},
\end{equation}
where $\bm{u}'$ is the velocity field \emph{without} the shear flow. Thus,
$\overline{U}$ has the effect of Doppler shifting the time derivative into the
frame comoving with the mean flow. If $\overline{U}$ is roughly constant, then
the behavior of a linear plane-wave perturbation satisfies the modified
dispersion relation
\begin{equation}
    \p{\omega - \overline{U}k_x}^2 =
        \frac{N^2k_{x}^2}{k_{x}^2 + k_{z}^2 + \p{2H}^{-2}}.
        \label{eq:disp_rel_U}
\end{equation}
This is just Eq.~\eqref{eq:disp_rel} with $\omega \to \omega - \overline{U}
k_x$. It is apparent that if $\overline{U} = \overline{U}_c$, where
\begin{equation}
    \overline{U}_c \equiv \frac{\omega}{k_x},\label{eq:u_crit}
\end{equation}
then the dispersion relation is singular and the linear solution breaks down.
Physically, this corresponds to the Doppler-shifted frequency of the IGW being
zero. Anywhere $\overline{U} = \overline{U}_c$ is called a \emph{critical
layer}.

The behavior of an IGW incident upon a critical layer was first studied in the
inviscid, linear regime in \citep{booker_bretherton}, which found nearly
complete absorption of the IGW\@. The amplitude
reflection and transmission coefficients are given by
\begin{align}
    \mathcal{R} &= \exp\p{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= \exp\p{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
        \label{eq:crit_coeffs}
\end{align}
where $\mathrm{Ri}$ is the local Richardson number evaluated at the critical
layer height $z_c$:
\begin{equation}
    \mathrm{Ri} \equiv \at{\frac{N^2}{\p{\pdil{\overline{U}}{z}}^2}}_{z_c}.
        \label{eq:ri_def}
\end{equation}
In the $\mathrm{Ri} \gg 1$ limit, $\mathcal{R}, \mathcal{T} \ll 1$ and the
incident wave is almost completely absorbed. This result also applies to viscous
fluids \citep{hazel}. However, weakly nonlinear theory \citep{brown_stewartson}
and numerical simulations \citep{winters1994} suggest that nonlinear effects may
significantly enhance reflection and transmission.
